Inflammation, SARS-CoV-2 infection and mental health disorders impact on systemic serotonin and dopamine metabolism

### 2023-05-23 paper/markdown/rebuttal_letter.Rmd

---
title: "Inflammation, SARS-CoV-2 infection and mental health disorders impact on systemic serotonin and dopamine metabolism"
subtitle: "Rebuttal Letter"
author: "PSY Team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: infl_biblio.bib
csl: frontiers-in-immunology.csl
header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      dev = "png", dpi = 600, 
                      tab.cap.pre = 'Supplementary Table ', tab.cap.sep = " ", 
                      tab.cap.style = 'Table Heading')

set_flextable_defaults(font.family = 'Cambria', font.size = 10)

```

\newpage

# Editor

<span custom-style = 'reviewer'>Thank you for the timely and well written piece. 
I have received three reviews from exceptional scholars in the field. 
As you can see, there are many strengths to the paper that they note and the publication is timely. 
However, there are also several concerns raised and I am not entirely sure they can all be sufficiently addressed to warrant publication. 
However, I think it is worth a shot. 
Please be attentive to reviewer concerns for a resubmission.</span>

Thank you for appreciation of our manuscript and giving us a chance to address Reviewers' issues. 
Rewrote the paper and put efforts to incorporate the critic in the text. 
In particular, we defined clearly the analysis endpoints, and decided to present our story in a 'top - bottom' approach. 
In brief, we employed the most robust and comprehensive analytic technique, multi-parameter modeling, to identify the most relevant explanatory factors affecting systemic serotonin and dopamine metabolism. 
Subsequently, we explore such factors in more detail by correlation and network analysis, and canonical statistical tests. 
Finally, we improved quality of language and Discussion. 
Please see below for details. 
We hope that, our revised manuscript is now suitable for publication in _Psychoneuroendocrinology_.

\newpage

# Reviewer 1

## Issue 1

<span custom-style = 'reviewer'>The study assesses the relationship between psychological variables, inflammatory markers, and concentrations of aminoacids in individuals after SARS-CoV-2 infection compared to uninfected controls in two distinct cohorts (SIMMUN; INCOV). 
Moreover, the difference between the control group and different states after SARS-CoV-2 infection (acute, sub-acute, recovery) are compared to the control group within the longitudinal INCOV study cohort.
In addition, the study aims to test the hypothesis that Phenylalanine, Tyrosine, Tryptophan, Kynurenine, and their respective ratios, are predicted by psychological variables inflammation, and SARS-CoV-2 infection-related variables (infection yes vs no, Anti-Antibody titers).
First, it should be mentioned that the graphs are of very high quality and aesthetics and summarize the results very well. 
In addition, the use of a very sophisticated analysis (multiparametric linear modeling) and the detailed presentation of this analysis, and the R packages used are highly creditable. 
However, several weaknesses, including, biochemically wrong assumptions regarding the assessed parameters, lack of transparency regarding the pre-processing of the data, and measurement comparability between the assessed cohorts limit its significance and impact. 
The quality of the paper also suffers from a lack of references, scientifically inadequate language, and multiple spelling and grammatical errors. Because of the many issues that arise, I would not recommend the manuscript for publication in _Psychoneuroendocrinology_.</span>

We thank the Reviewer for the careful lecture of our manuscript and valuable feedback. 
In the revised manuscript, we did our best to address the weaknesses pointed out be the Reviewer.
In order to improve the clarity of our research aims and to facilitate interpretation of the results, we re-formatted the analysis pipeline and the manuscript in the following way:

* we streamlined the __Introduction__ section. Here, we describe briefly the serotonin and dopamine metabolism, and the competitor kynurenine pathway, functions of those neurotransmitter in the central nervous system and impact on serotonin, dopamine on mental health. We outline the role of inflammation and mutual relationship between tryptophan metabolism and mental health. Furthermore, we review literature evidence for the SARS-CoV-2/COVID-19 - neurotransmitter connex.

* In __Introduction__, we also define clearly the 'evidence gap' and our research aims in the last paragraph of the section.

* In  __Methods/Analysis endpoints__ we define unequivocally the outcomes, i.e. plasma levels of metabolites involved in serotonin and dopamine turnover, and explanatory factors, i.e. demographic, clinical, inflammatory, psychometric and SARS-CoV-2-associated variables.

* We restrict the analyzed SIMMUN and INCOV cohorts to participants with the complete set of outcome and explanatory variables (__Figure 1__, __Supplementary Table S1__ and __S3__). By this change, results of particular analyses can be more easily compared with each other. Additionally, to investigate if and how exclusion of participant affected the SIMMUN cohort characteristic and if it could be regarded as a source of bias, we compared th excluded and analyzed individuals as presented in __Supplementary Table S2__ and __Results/Characteristic of the study cohorts__.

* We excluded CRP and IL6 from analyses presented in the revised manuscript. Those markers were found to be slightly above their reference ranges (0.5 mg/dL and 7 pg/mL) solely in 
`r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$crp_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% and 
`r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$il6_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% of SIMMUN individuals, respectively. They are hence unlikely to play a clinically relevant role in regulation of TRP, KYN, KYN/TRP, PHE, TYR and PHE/TYR, in spite of statistically significant correlations presented in the initial manuscript. We argument this decision in __Supplementary Methods__ of the revised manuscript.

* We provided information on all analyzed variables in the SIMMUN (__Supplementary Table S1__) and INCOV cohort (__Supplementary Table S3__) with their format, units and, if applicable, normality-stabilizing transformations.

* In order to improve clarity of our results, we decided to present our results following a 'top - bottom' scheme. In brief, both in the SIMMUN and the INCOV cohort, we employ multi-parameter modeling (linear regression or robust linear regression), to identify key explanatory factors significantly affecting the outcome. Subsequently, we investigate those significant predictions in more detail by 'univariable' correlation analysis and statistical hypothesis testing. This analysis strategy was pursued in our recent papers as well [@Hufner2022; @Hufner2023].

## Issue 2

<span custom-style = 'reviewer'>__Introduction:__ In the introduction the term "psychoneuroimmunological mechanisms" is mentioned twice. 
I would appreciate it if at least some of the mechanisms which are relevant to the paper are specified.</span>

As outlined in __Issue 1__, we re-wrote thoroughly the __Introduction__ section. 
Following discussions in the core author team, we decided no to replace the general and a bit obscure term 'psychoneuroimmunological' with more direct formulations referring to the prime role of inflammation in modulating serotonin and dopamine biosynthesis. 
For this reason, we describe now explicitly the effects of inflammatory cytokines on IDO- and PAH-mediated regulation of serotonin and dopamine biosynthesis and changed the manuscript title to "Inflammation, SARS-CoV-2 infection and mental health disorders impact on systemic serotonin and dopamine metabolism". 
In fact as discussed below and in the revised manuscript text, the question how systemic biosynthesis of serotonin and dopamine, involving primarily mesenteric organs and the liver, relates to the neurotransmitter availability in the brain is vigorously disputed in literature [@Goldstein1999; @Moncrieff2022; @Marx2020].

## Issue 3

<span custom-style = 'reviewer'>__Introduction:__ Some statements made in the introduction are without reference, very general, and the conclusions weakly justified. 
For instance, TRP, KYN, and KYN/TRP are listed as markers of serotonin availability. 
However, the available evidence for the relationship between increased degradation of TRP to KYN and its effects on serotonin availability is controversial as only 1-2% of tryptophan metabolism goes into the serotonin pathway (https://doi.org/10.3390/metabo12060514). 
In addition, line 60 mentions a depressiogenic and an anxiogenic effect of QUIN, and an anxiety- and depression-amplifying effect of kynurenine and catabolites is stated in line 9 (page 3). 
Should this statement refer to existing cross-sectional findings (e.g., higher QUIN in patients with MDD compared to healthy controls) it is very generalized and not sufficiently justified. 
If the statement refers to evidence that suggests a causal relationship, a reference is urgently needed. </span>

We agree with the Reviewer, that __Introduction__ needs to be more strict and requires additional references - please refer to the revised text. 
Concerning your more specific issues: we agree that the great majority majority of tryptophan is catabolized via the KYN and indole pathways and only a marginal fraction is diverted to serotonin synthesis [@Teunis2022; @Lukic2022], and for both serotonin and dopamine, the systemic metabolism taking place in liver, mesenteric organs and vasculature may not reflect the situation in the central nervous system [@Brown2021; @Marx2020; @Schwarcz2012; @Moncrieff2022; @Goldstein1999]. 
We take up this controversy in __Introduction__ and __Discussion__. 
Additionally, we addressed the question how TRP levels may affect systemic serotonin availability, we modeled circulating serotonin levels in the INCOV cohort [@Su2022] as a function of its precursors and competitor metabolites (TRP, KYN, QUIN), timepoint of SARS-CoV-2 infection, inflammatory cytokines and age. 
As presented in __Figure 3__, we could indeed find a significant positive association of plasma TRP and serotonin in multi-parameter modeling, which is expected to be independent of age, inflammation and infection time course. 
In a more detailed correlation and network analysis of the INCOV dataset (__Figure 5A__), we could observe a moderate-strength correlation of circulating serotonin and TRP levels in acute and sub-acute infection. 
This suggests that systemic serotonin synthesis may be limited by blood TRP availability to some extent during SARS-CoV-2-dependent inflammation, which was characterized by profoundly decreased TRP and elevated KYN levels (__Figure 4__). 
In the revised __Introduction__ and __Discussion__, we also critically review the evidence on brain metabolism and effects of KYN, kynurenic acid and QUIN on neuronal signaling and mental health [@Brown2021; @Schwarcz2012], which is indeed limited.

## Issue 4

<span custom-style = 'reviewer'>__Introduction:__ The introduction sentence of the last paragraph in the introduction seems not adequate for the following hypotheses. 
It highlights that low-grade inflammation (as I understand it, in the case of this study IL-6 and CRP concentrations) is associated with mental health. 
However, the presented hypotheses use IL-6 and CRP and mental stress, anxiety, and depressive symptoms as predictors in their analysis (multi-parameter linear modeling) and TRP, KYN, KYN/TRP, PHE, TRY, PHE/TRY as dependent variables.</span>

We are sorry for this unclarity. 
In the revised manuscript, we clearly formulate the aims, and define outcomes and explanatory factors. 
Please refer to the last paragraph of __Introduction__, __Methods/Analysis endpoints__, __Supplementary Tables S1__ and __S3__. 
In the revised text, we additionally avoid the misleading term 'low-grade' inflammation. 
In fact, as described in __Issue 1__ and __Supplementary Methods__, only a minute subset of the SIMMUN study participants had CRP or IL-6 concentrations above the reference levels. 
Thus, biological effect of those two inflammatory markers is doubtful and we decided to exclude them from the revised analysis pipeline.

## Issue 5

<span custom-style = 'reviewer'>__Methods:__ while the SIMMUN cohort is described in detail, basic information about the INCOV cohort is missing and should be provided.</span>

We agree with the Reviewer that the INCOV cohort deserves a more thorough description. This is provided in the __Methods/Procedures__ section of the revised manuscript and in __Supplementary Methods__ and in references to the seminal papers [@Su2020; @Su2022].

## Issue 6

<span custom-style = 'reviewer'>__Methods:__ Before the statistical analysis, several transformations (log2, ln, and square root) were applied. 
However, it is not transparent how each parameter is transformed. I would appreciate it if this could be further specified.</span>

We provide now these information in __Methods__ and __Supplementary Methods__. 
Variable format, transformations and, if appropriate, stratification schemes are presented in __Supplementary Tables S1__ and __S3__.

## Issue 7

<span custom-style = 'reviewer'>Methods: Furthermore, when measured within clinical routine laboratory values of CRP and IL-6 are in many cases lower the limits of detection (e.g., 0.3 mg/ml). 
Was this the case in this study? 
If so, it would be useful for the replicability to know how the authors dealt with this issue.</span>

As specified now in __Methods/Procedures__, values beyond or at their detection limits were substituted with the lower or upper detection limit. 
CRP and IL6 were removed from the revised analysis pipeline for the reasons specified in __Issue 1__.

## Issue 8

<span custom-style = 'reviewer'>__Methods:__ Kruskal-Wallis test with Man-Whitney U tests are applied to assess the differences be-tween uninfected individuals and infected individuals at different time points. 
While the pairwise comparisons are adequate for this research question, the overall statistic of the Kruskal-Wallis test is not useful in this context, as it tests whether there is an over-all difference between independent groups. 
I would therefore suggest using a linear model with predefined contrasts (uninfected vs. acute; uninfected vs. sub-acute; uninfected vs. recovery) inclusive control for multiple tests. 
If the assumptions of linear regression are not fulfilled, robust methods can be used (e.g., R-package: MASS; https://www.rdocumentation.org/packages/MASS/versions/7.3-58.2/topics/rlm). 
Additionally, the correlations between the assessed dependent variables seem rather high, why controlling for multiple comparisons using Benjamini & Yekutieli procedure (2001) or even robust MANOVAS (see: https://cran.r-project.org/web/packages/MANOVA.RM/vignettes/Introduction_to_MANOVA.RM.html) could be also necessary. Additionally, the terms "acute", "sub-acute and "recovery" should be described in the notes of __Figure 3__.</span>

We are grateful for the hints. 
In the revised manuscript, we applied robust linear regression with the MM algorithm and Huber's psi function [@Huber2011; @Ripley2022] for multi-parameter modeling (__Figure 3__) and time-course modeling of the INCOV dataset (__Figure 4__). 
Additionally, we specify in the figure legends of the INCOV results the median time interval between diagnosis and sampling for the acute, sub-acute and recovery timepoints. 
As suggested, we applied multiple testing adjustment throughout the analysis pipeline. 
Except for multi-parameter modeling, p values were adjusted separately for each analysis step (e.g. correlation analysis, time-course modeling) with the false discovery rate method known also as 'Benjamini-Hochberg' [@Benjamini1995].

## Issue 9

<span custom-style = 'reviewer'>__Psychometric assessment:__ In the description of the psychometric assessment, the source of the HADS is missing as well as a calculation of the reliability of the used questionnaires for the SIMMUN study cohort.</span>

We apologize for the missing references and cite the seminal HADS paper and a report dealing with the clinically relevant HADS cutoffs [@Bjelland2002; @Zigmond1983] in the revised text. 
We also checked consistency of the PSS-4 and HADS psychometric tools employed in our report. 
Since the HADS scales, and in particular, the HADS depression tool were found to be non-tau-equivalent by factors analysis (__Supplementary Figure S1A__), we decided use McDonald's $\omega$ as a consistency measure instead of the more popular Cronbach's $\alpha$ [@Revelle2022; @McDonald1999]. 
The PSS-4 tool demonstrated a good and the HADS scales an excellent consistency, as presented in __Supplementary Figure S1B__.

## Issue 10

<span custom-style = 'reviewer'>__Results:__ First, it would be useful to distinguish between cohorts in the descriptive characterization of cohorts. 
Furthermore, concrete numbers should be given instead of rough descriptions (e.g., "about half" (page 8, line 19)). 
Second, assessing the effect as "dramatic" is not appropriate in a scientific context. 
If at all, existing assessments based on effect sizes can be made using generally accepted definitions (e.g., according to Cohen).</span>

We concur with the Reviewer, that a comparison between the cohorts is useful for interpretation of the results. 
We present significant results of the comparison between the SIMMUN and INCOV cohort in __Supplementary Table S4__ and elaborate on them in __Results/Characteristic of the study cohorts__. 
The collectives differed in age and sex distribution, rates of SARS-CoV-2 infection and infection severity. 
In particular, most INCOV study participants experienced moderate-to-critical COVID-19 and required hospitalization. 
This may explain the way stronger effects of inflammation and infection timepoint on inflammatory cytokines and the metabolites of interest, which we discuss in __Discussion__ and __Limitations__.
In the revised text, Figures and Tables we also present effect size metrics for statistical hypothesis tests, correlation and modeling. 
We interpret them as proposed in the excellent handbook by Jacob Cohen [@Cohen2013] (__Supplementary Methods__).

## Issue 11

<span custom-style = 'reviewer'>__Results:__ In general, numbers (test statistics, p-values) are missing in the results part and should be added. Additionally, effect sizes should be presented.</span>

We present now p values and effect sizes or model $\beta$ with 95% confidence intervals [@Cohen2013], as appropriate, for the most important findings in Abstract and Results. 
Yet, for the sake of readability, we were not able to provide significance and effect size information for all effects. 
In particular, we do not specify p values and effect sizes for non-significant effects and state instead clearly that the effect missed statistical significance. 
We abstained from referencing test statistics such as $\chi^2$ or $t$, since they depend on the number of complete observations and hence do not deliver reliable information on the effect size.

## Issue 12

<span custom-style = 'reviewer'>__Results:__ "Tends to be higher"/ "Slightly more" is scientifically incorrect. 
Please replace the terms and add the test statistic and p-value</span>

Following your suggestion, we avoid this wording in the revised manuscript. 
Please refer to __Issue 11__ for details.

## Issue 13

<span custom-style = 'reviewer'>__Results:__ The terminology of strata (page 10, line 10) can be misleading as no stratification was performed. Probably it would be better to use the term group or subgroup.</span>

Following your suggestion, we avoid using the term 'strata' in the revised text. 
We use group, subgroup or subset instead.

## Issue 14

### 2023-05-23 paper/markdown/main_manuscript.Rmd
Manuscript and Rebuttal Letter

---
title: "Persistent low-grade inflammation, SARS-CoV-2 infection and mental health deterioration signs impact on systemic levels of aminoacid neurotransmitter precursors"
subtitle: "Manuscript"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

@@

## Data and code availability

Anonymized patient data will be made available upon request to the corresponding author. 
The INCOV cohort data are publicly available at https://data.mendeley.com/datasets/96v329bg7g/1.
The analysis pipeline code is available at https://github.com/PiotrTymoszuk/simmun_revision.

>>>>

---
title: "Inflammation, SARS-CoV-2 infection and mental health disorders impact on systemic serotonin and dopamine metabolism"
subtitle: "Manuscript"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

author:
  - Katharina Hüfner:
      institute: [psych2]
      email: katharina.huefner@tirol-kliniken.at
      correspondence: true
  - Sophia Vedova:
      institite: [psych2]
  - Piotr Tymoszuk:
      institute: [daas]
  - Philipp Nelles:
      institute: [psych2]
  - Markus Anliker: 
      institute: [zimcl]
  - Tobias Bruckner: 
      institute: [psych2]
  - Eberhard Deisenhammer: 
      institute: [psych2, psych1]
  - Jonas Egeter: 
      institute: [psych2]
  - Alexander Egger: 
      institute: [zimcl]
  - Matyas Galffy: 
      institute: [psych2]
  - Johannes Giesinger: 
      institute: [psych2]
  - Jens Lehmann: 
      institute: [psych1]
  - Maria Oberhammer: 
      institute: [psych1]
  - Joachim Rockenschaub: 
      institute: [psych2]
  - Magdalena Sacher: 
      institute: [surg]
  - Bernhard Holzner: 
      institute: [psych2]
  - Johanna M Gostner: 
      institute: [bio]
  - Barbara Sperner-Unterweger: 
      institute: [psych2]

institute:
  - psych2: Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, University Hospital for Psychiatry II, Medical University of Innsbruck, Anichstrasse 35, Innsbruck, Austria
  - daas: Data Analytics As a Service Tirol, Innsbruck, Austria
  - zimcl: Central Institute of Medical and Chemical Laboratory Diagnostics, University Hospital Innsbruck, Anichstrasse 35, Innsbruck, Austria
  - psych1: Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, University Hospital for Psychiatry I, Medical University of Innsbruck, Anichstrasse 35, Innsbruck, Austria
  - surg: Department of Visceral, Transplant and Thoracic Surgery, Center of Operative Medicine, Medical University of Innsbruck, Innsbruck, Austria
  - bio:  Institute of Medical Biochemistry, Biocenter, Medical University of Innsbruck, Innrain 80/82, Innsbruck, Austria

@@

# Keywords

SARS-CoV-2, inflammation, serotonin, dopamine, tryptophan, kynurenine, psychiatric disorders

# Abstract

__Background:__ Serotonin and dopamine metabolism poses a junction of mental and physical health. 
We aimed to identify demographic, clinical, psychometric and SARS-CoV-2 infection-related factors affecting systemic serotonin and dopamine turnover.

__Methods:__ The cross-sectional SIMMUN (n = `r nrow(stigma$analysis_tbl)`, Austria) and longitudinal INCOV cohort (n = `r  length(unique(incov$complete_ids))`, Su et al. 2022) were investigated. 
Explanatory variables encompassed age, sex, clinical and inflammatory variables, SARS-CoV-2 infection, signs of anxiety and depression (HADS), and mental stress scoring (PSS-4). 
Their effects on serotonin biosynthesis markers (serotonin, tryptophan) and competitor kynurenine pathway products (kynurenine, quinolinate), and dopamine availability readouts (dopamine sulfate, phenylalanine, tyrosine) were assessed by linear modeling, correlation analysis and two-tailed T tests.

__Results:__ In the SIMMUN cohort, the inflammatory marker neopterin, SARS-CoV-2 infection, age, depression symptoms and mental stress were positively associated with catabolism of the serotonin precursor tryptophan to kynurenine. 
Inflammation and age were also suppressed the first step of dopamine biosynthesis, phenylalanine - tyrosine conversion. 
In the INCOV cohort, inflammation was associated with lowered serotonin (IL6: $\beta$ = `r incov_neuro$inference$serotonin %>% filter(variable == 'IL6_INF') %>% mutate(beta_lab = paste0(signif(estimate, 2), ' [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), ']')) %>% .$beta_lab`) and the dopamine metabolite, dopamine sulfate (interferon-gamma: $\beta$ = `r incov_neuro$inference$dopamine %>% filter(variable == 'IFNG_INF') %>% mutate(beta_lab = paste0(signif(estimate, 2), ' [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), ']')) %>% .$beta_lab`). 
Serotonin ($\beta$ = `r incov_neuro$inference$serotonin %>% filter(variable == 'timepoint', level == 'recovery') %>% mutate(beta_lab = paste0(signif(estimate, 2), ' [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), ']')) %>% .$beta_lab`) 
and dopamine sulfate ($\beta$ = `r incov_neuro$inference$dopamine %>% filter(variable == 'timepoint', level == 'recovery') %>% mutate(beta_lab = paste0(signif(estimate, 2), ' [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), ']')) %>% .$beta_lab`) were significantly increased during SARS-CoV-2 infection recovery.

__Conclusion:__ 
Inflammation limits systemic serotonin and dopamine by activating the competitor kynurenine pathway and inhibiting the phenylalanine - tyrosine conversion, respectively, and may link SARS-CoV-2 infection with mental health. 
Advanced age and psychiatric disorders can additionally suppress neurotransmitter synthesis.

# Introduction

Physical conditions such as infections, trauma, frailty, autoimmune illness or malignancy are often accompanied by sings of depression, anxiety, metal stress and other psychiatric disorders [@Sorgdrager2019; @Dantzer2008; @Hufner2019; @Hufner2020; @Maes2012; @Geisler2013; @Haroon2012; @Hufner2015]. 
Imbalance of two depressiolytic and anxiolytic neurotransmitters, serotonin and dopamine, elicited by inflammation as proposed as a connex of physical pathology and mental health deficits [@Dantzer2008; @Brown2021; @Schwarcz2012]. 
Serotonin is synthesized from the essential aminoacid tryptophan (TRP) via reactions catalyzed by tryptophan hydroxylase (TPH) and aromatic L-amino acid decarboxylase. 
Yet the great majority of TRP is by the so called kynurenine (KYN) pathway, whose first step is the breakdown of TRP to KYN catalyzed by indoleamine 2,3-dioxygenase (IDO) [@Badawy2017; @Lukic2022]. 
Expression of an inducible and highly active form of IDO, IDO1, is strongly stimulated by a plethora of cytokines, pathogen products and other inflammatory stimuli [@Robinson2003; @Dai1990; @Litzenburger2014; @Badawy2017]. 
During systemic or central nervous system-localized inflammation, TRP is depleted by high IDO1 activity, which, in turn, suppresses serotonin biosynthesis [@Brown2021; @Schwarcz2012; @Vancassel2018; @Haroon2012]. 
Consequently, elevated blood levels of KYN pathway metabolites, decreased TRP and serotonin were reported for numerous physical conditions and associated with symptoms of psychiatric deficits [@Brown2021; @Marx2020; @Hunt2020; @Fellendorf2022; @Hufner2015; @Hufner2021; @Hufner2019; @Paul-Savoie2011; @Capuron2011]. 
Dopamine synthesis involves the conversion of another essential amino acid phenylalanine (PHE) to tyrosine (TYR) catalyzed by phenylalanine hydroxylase (PAH) followed by hydroxylation of TYR by tyrosine 5-hydroxylase (TH) [@Meiser2013]. 
Both PAH and TH as well as the serotonin-synthesizing enzyme TPH rely on tetrahydrobiopterin (BH4) as an essential cofactor. 
During inflammation, BH4 is depleted by oxidation to its inactive derivative and gets consumed in a competitor reaction of nitric oxide formation. 
As a result of such functional BH4 deficiency, biosynthesis of dopamine and serotonin is further suppressed [@Geisler2013; @Fanet2021; @Vancassel2018]. 
Accordingly, imbalance of PHE and TYR levels suggestive of systemic dopamine deficiency was reported in cancer, infections and inflammatory conditions and associated with depression and anxiety [@Hufner2015; @Neurauter2008; @Geisler2013; @Fanet2021; @Vancassel2018].

SARS-CoV-2 virus is the causal pathogen of coronavirus disease 2019 (COVID-19). 
Apart from sustained physical disability, psychiatric disorders in millions of SARS-CoV-2 patient amount to the persistent burden of the COVID-19 pandemic [@Huang2021; @Al-Aly2021; @Hufner2022; @Sahanic2023; @Staudt2022]. 
Of note, high activity of the KYN pathway and signs of inflammatory BH4 deficiency manifesting by distorted PHE and TYR levels were reported for COVID-19 patients and associated with inflammation and disease severity [@Gietl2023; @Lionetto2021; @Santiago-Mujika2022; @Luporini2021; @Bizjak2022; @Dewulf2022; @Saito2022]. 
Furthermore, reduced systemic availability of serotonin and dopamine was postulated to contribute to mental health deficits during SARS-CoV-2 infection recovery [@Bower2022; @Kucukkarapinar2022; @Matits2023]

Still, the impact of inflammation on systemic metabolism of serotonin and dopamine availability in acute SARS-CoV-2 infection and recovery, along with its mutual interaction with psychiatric disorders is still incompletely characterized. 
To address that, we explored effects of demographic and clinical factors, inflammation, SARS-CoV-2 infection, anti-SARS-CoV-2 humoral response and psychiatric disorder symptoms on readouts of systemic serotonin and dopamine availability and kynurenine pathway activity in two cross-sectional cohorts of SARS-CoV-2 patients: the SIMMUN cohort and the published longitudinal INCOV study [@Su2022].

# Materials and Methods

## Ethics statement

The SIMMUN study was conducted in accordance with the Declaration of Helsinki and European Data Policy. 
All participants gave written informed consent prior to enrollment. 
Participants' data were stored and analyzed in anonymized form.
This study was approved by the ethics committee of the Medical University Innsbruck (Austria, approval number: 1132/2020). 
No approval by the ethics committee was required for analysis of the published INCOV dataset [@Su2022].

## Study cohorts

Details on study cohorts, procedures and analyses are provided in __Supplementary Methods__.

### SIMMUN study

Individuals tested for SARS-CoV-2 via PCR at the University Hospital of Innsbruck and patients of the University Clinic for Psychiatry I and II (Innsbruck, Austria) undergoing routine SARS-CoV-2 PCR were invited to participate. 
The study was conducted between June 2020 and May 2021. 
The inclusion criteria were age of 18 - 70 years, proficiency in German, residence in the study region (Tyrol, Austria), and a SARS-CoV-2 PCR test conducted at the study site. 
The exclusion criteria were active SARS-CoV-2 infection (< 14 days after diagnosis), pregnancy, active malignancy, organ transplantation, surgery in the past 3 months, inflammatory illness and oral corticosteroid treatment. 
The analysis inclusion criterion was the complete study variable dataset (__Figure \@ref(fig:fig-inclusion)__, __Supplementary Table S1__). 
Significant differences between the analyzed and excluded participants are listed in __Supplementary Table S2__.

### INCOV study

Proteome and metabolome data and clinical information for the INCOV cohort were extracted from supplementary tables accompanying the report by Su and colleagues [@Su2022]. 
In the current analysis, samples obtained for individuals with the complete age information and complete study variable dataset (__Figure 1__, __Table \@ref(tab:tab-incov-cohort)__, __Supplementary Table S3__).

## Procedures

### SIMMUN study

The SIMMUN study data were gathered during a single on-site study visit including a general medical assessment, supervised completion of self-rating questionnaire and a blood sample collection (__Supplementary Table S1__). 
The median time interval between the study visit and SARS-CoV-2 PCR was 139 days (interquartile range: 119 - 157).

Demographic and clinical variables: age, sex, body mass index, professionally diagnosed psychiatric disorders, self-reported physical disorders, smoking and alcohol consumption, result and date of the SARS-CoV-2 PCR test were surveyed during the study visit or extracted from electronic patient records. 
Inflammatory markers: plasma neopterin concentrations (NEO) and neutrophil - lymphocyte ratio (NLR) were determined by the certified clinical routine laboratory at the University Hospital of Innsbruck. 
Plasma concentrations of tryptophan (TRP), kynurenine (KYN), phenylalanine (PHE) and tyrosine (TYR) were determined by high-performance liquid chromatography, and the KYN/TRP and PHE/TYR ratios calculated [@Capuron2011; @Widner1997; @Neurauter2008]. 
Plasma titer of immunoglobulin gamma against receptor binding domain S1/S2 protein (anti-RBD IgG) were quantified by ELISA [@Deisenhammer2021]. 
Laboratory measurements at and beyond the detection limits were substituted with the lower or upper detection limit value, respectively. 
Mental stress was scored with the 4-item perceived stress scale (PSS-4) [@Cohen1983]. 
Anxiety and depression signs were scored with the hospital anxiety and depression scale (HADS) including 7 items for anxiety and 7 items for depression [@Zigmond1983]. 
Clinically relevant signs of anxiety or depression were identified with the cutoff of $\geq$ 8 points [@Bjelland2002; @Zigmond1983].

In order to improve normality of some numeric study variables prior to linear modeling and statistical hypothesis testing with parametric tools, logarithm or square root transformations were applied (__Supplementary Table S1__)

### INCOV cohort

Plasma proteomes and metabolomes in the INCOV cohort were measured by proximity extension assay (Olink, Sweden) and ultra-high-performance liquid chromatography/tandem accurate mass spectrometry (Metabolon, USA) [@Su2022; @Su2020]. 
Normalized, age- and sex-adjusted, log~2~-transformed plasma concentrations of metabolites and cytokines, and clinical information were extracted from [supplementary tables](https://data.mendeley.com/datasets/96v329bg7g/1) of the report by Su et al. [@Su2022]. 
Metabolites implicated in systemic serotonin availability (serotonin, TRP), KYN pathway activity (KYN, quinolinate [QUIN]) and dopamine availability (PHE, TYR and dopamine 3-O-sulfate [DA sulfate]), cytokine markers of inflammation (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) and age were analyzed (__Supplementary Table S3__).

Plasma features were sampled in uninfected controls and SARS-CoV-2 individuals during 
acute (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['acute']`) and 
sub-acute (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['sub-acute']`) infection, and recovery (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['recovery']` days after diagnosis of SARS-CoV-2 infection via PCR) (__Supplementary Table S3__).

## Analysis endpoints

Our analysis pursued two endpoints. 
The first endpoint was to determine demographic, clinical, psychometric, inflammation- and SARS-CoV-2-related factors influencing plasma levels of serotonin and dopamine precursors and KYN pathway metabolites (TRP, KYN, KYN/TRP, PHE, TYR, PHE/TYR). 
This endpoint was addressed by multi-parameter modeling in the SIMMUN cohort. 
The second endpoint was to investigate how more direct readouts of systemic serotonin and dopamine availability, plasma serotonin [@Paul-Savoie2011; @Sun2020; @Tateishi2021; @Holck2019] and DA sulfate [@Meiser2013; @Hashizume1987; @Goldstein1999], are influenced by age, inflammatory cytokines, timepoint of the SARS-CoV-2 infection, neurotransmitter precursors (TRP, PHE, TYR) and KYN pathway products (KYN, QUIN). 
This endpoint was addressed by multi-parameter robust linear modeling, time course modeling and correlation analysis in the INCOV cohort (__Figure \@ref(fig:fig-inclusion)__).

## Statistical analysis

R version 4.2.3 was used for statistical analysis.

Numeric variables were presented as medians with interquartile ranges and ranges. 
Categorical variables are presented as percentages and counts within the complete observation set. 
Distribution normality and variance homogeneity was assessed by Shapiro-Wilk and Levene test, respectively. 
In the SIMMUN cohort, non-normally distributed numeric variables were logarithm- or square root-transformed prior to modeling and parametric tests (__Supplementary Table S1__). 
Since most of the INCOV study variables were non-normally distributed, robust linear modeling and non-parametric testing were employed. 
Since the tau-equivalence assumption investigated by factor analysis for the HADS scales was violated, consistency of psychometric tools was investigated by global McDonald's $\omega$ [@McDonald1999; @Bartlett1937; @Revelle2022]. 
Except for multi-parameter modeling, p values were corrected for multiple testing with the false discovery rate method [@Benjamini1995] separately for each analysis task. 
Effects with p < 0.05 were considered significant.

Correlation was assessed by Spearman's rank or Pearson test. 
Significance of comparisons of numeric variables between two groups was determined by Mann-Whitney test with r effect size statistic or by two-tailed T test with Cohen's d effect size metric. 
Comparisons of categorical variable distributions were evaluated by $\chi^2$ test with Cramer V effect size statistic. 
Correlation was assessed by Pearson's or Spearman's rank test [@Cohen2013; @Kassambara2021].
